A primer to numerical simulations: The perihelion motion of Mercury

### 2017-12-20 tex/MercuryNumerical.tex
added new figures and beta parameter to scripts

To allow for a movement of the perihelion we need to modify the force that appears in Eq.~(\ref{eq:newton}).
This modification should be such that it depends on $r$ and it vanishes for $r\to \infty$.
Based on the discussion presented above we may multiply the right hand side with a factor
$(1+\alpha(r_S/r))$, where $\alpha$ denotes some dimensionless parameter. 
However, besides $r_S$, which is the characteristic length scale for the gravitational potential
of the sun, one can define one more parameter, which is characteristic for the dynamics of mercury, namely
\begin{equation}
h=\frac{\vec L\,^2}{m^2c^2}= \frac{(\vec r\times \dot{\vec r} \, )^2}{c^2} \ ,
\end{equation}
where $\vec L$ denotes the angular momentum, which is a constant of motion for central potentials.
It is easily verified that $h$ carries dimensions of a length squared.  Therefore we may use  the following ansatz
for the modified potential
\begin{equation}
\ddot{\vec r} = - \frac{c^2}{2}\left(\frac{r_S}{r^2}\right)\left(1+\alpha\frac{r_S}{r}+\beta\frac{h}{r^2}\right) \, \frac{\vec{r}}{r} \ .
\label{eq:newton_art}
\end{equation}
For the system at hand $h$ may be estimated from the parameters of mercury at its perihelion ---  $\abs{\vec v_M(0)} = 59$km/s and 
$ r_{MS}=\abs{\vec r_{MS}(0)} = 46 \cdot 10^6$km~\cite{MercuryFactSheet}, where both $\vec r$ and $\dot{\vec r}$ are perpendicular to each other. 
We thus find that $(h/ r_{MS}^2) = 4\times10^{-6}$. Furthermore $(r_S/r_{MS})=6\times 10^{-6}$. 
Therefore one may expect from both correction terms an effect similar in size. Furthermore, any term higher order in either $h$ or $r_S$
should be suppressed by additional six orders of magnitude and thus can be safely neglected.
The parameters $\alpha$ and $\beta$ could now either be extracted from a fit to data or be calculated from the underlying
theory --- here General relativity. However, since we have only a single number to study, the angular motion of the perihelion,
we can only fix either $\alpha$ or $\beta$.

@@

In \python{}, this reads
\begin{lstlisting}
# Definition of parameters
rM0 = 4.6    # in units of R0
vM0 = 0.51   # in units of R0/T0
c_a = 1.01   # in units of R0/T0**2
TM  = 88.    # in units of T0
rS  = 3.e-7  # in units of R0
\end{lstlisting}
So far we only fixed the length of the vectors.
Next, we set up the initial directions, which will describe the motion in the two-dimensional plane.  We will build on the existing \python{} module \texttt{VPython}, which provides an implementation for treating vectors as well as their visualization.
The first object of interest is a \texttt{vector}, which takes three-dimensional coordinates as its input.
Because of our choice of initial conditions (we picked the initial vectors in the perihelion), the velocity of mercury is perpendicular to the vector which connects mercury and sun (see figure \ref{fig:sun_merc}):
\begin{lstlisting}
# Import the class vector from vpythons module visual
from vpython import vector
# Initialize distance and velocity vectors
vec_rM0 = vector(0, rM0, 0)
vec_vM0 = vector(vM0, 0, 0)
\end{lstlisting}
According to Eq.~(\ref{eq:newton}), the force which acts on mercury changes the velocity of mercury which eventually changes the position.
However, we first need to fix the time step $\Delta t$ and use the estimate of Eq.~(\ref{eq:check}) as a guidance
\begin{lstlisting}
# Definition of the time step
dt = 2 * vM0 / c_a / 20
\end{lstlisting}
Here the factor $1/20$ makes sure that $\Delta t$ is indeed consistent with Eq.~(\ref{eq:check}).
Now we are in the position to calculate location and velocity of the planet at $t_0+\Delta t$ using the following equations\footnote{%
	Note, that there are different notions for numerically integrating differential equations with different accuracys.
	Thus, the ordering and exact equations for updating $a_{MS}$, $v_M$ and $r_M$ are not unique, but all correct prescriptions lead to the same result for sufficiently small $\Delta t$.
	For this particular problem, the following definition is in good balance between numerically stable and inexpensive to compute.
}%
\begin{lstlisting}
# Compute the strength of the acceleration
aMS = c_a * ( 1 + alpha * rS / vec_rM_old.mag  ) / vec_rM_old.mag**2
# Multiply by the direction
vec_aMS = - aMS * ( vec_rM_old / vec_rM_old.mag )
# Update velocity vector
vec_vM_new = vec_vM_old + vec_aMS * dt
# Update position vector
vec_rM_new = vec_rM_old + vec_vM_new * dt
\end{lstlisting}
Note basic vector operations are already implemented in the predefined \texttt{vector} class.
The difference and sum of two vectors, or the scalar vector multiplication return vectors themselves.
Also the magnitude of a vector --- \texttt{vector.mag} --- is a property of the vector and can be easily extracted.

It is handy to use \texttt{Python}s \texttt{function}s to embed repeating structures ("DRY" --- Don't Repeat Yourself)
\begin{lstlisting}
# Define the coordinate and velocity update function
def evolve_mercury(vec_rM_old, vec_vM_old, alpha):
	# Compute the strength of the acceleration
	aMS = c_a * ( 1 + alpha * rS / vec_rM_old.mag  ) / vec_rM_old.mag**2
	# Multiply by the direction
	vec_aMS = - aMS * ( vec_rM_old / vec_rM_old.mag )
	# Update velocity vector
	vec_vM_new = vec_vM_old + vec_aMS * dt
	# Update position vector
	return vec_rM_new, vec_vM_new

# Call the function
vec_rM_new, vec_vM_new = evolve_mercury(vec_rM_old, vec_vM_old)
\end{lstlisting}
\python{}s syntax enforces a clean programming style: it is necessary that the body of the function is indented (by an arbitrary but consistent amount of \texttt{space}s or \texttt{tab}s) relative to the definition statement of the function.
Furthermore, \python{} is an Interpreter language.
Each line of the code is executed when the Interpreter passes it.
For this reason, all the variables defined before the function are global and thus known to the function.
The variables within the function (e.g. \texttt{aMS}, \texttt{vec\_aMS}, $\dots$) are local and do not exist beyond the scope of the function.

Finally, we can describe the evolution by a \texttt{while}-loop
\begin{lstlisting}%[float,floatplacement=H]
t     = 0
alpha = 0
# Execute the loop as long as t < 2*TM
while t < 2*TM:
	vec_rM, vec_vM = evolve_mercury(vec_rM, vec_vM, alpha)
	t = t + dt
\end{lstlisting}
where for the start we set the parameter $\alpha=0$ in order first study the properties of the pure 
$1/r^2$ force.
Note the required indent of the loop structure similar to the indent of a function.
In each iteration of the \texttt{while}-loop, the previous distance and velocity are used to compute the new values --- which directly overwrites the previous values and thus enter the next iteration.
The total runtime \texttt{2*TM} is the amount of "virtual" days the simulations should run.
To describe at least one full orbital period, this time needs to be larger than $T_M$.
With the previous choice for \texttt{dt}, this corresponds to roughly $N_T \approx 2 \cdot 10^3$ evolution steps.
Note that the exact time it takes for mercury complete a full revolution depends on the initial coordinates and velocities as well as the accuracy of the computation.
One might encourage students to analyze this in the beginning.

@@

Last but not least, instead of updating the object-unrelated vectors in the while loop, we update the object-related vectors which will eventually be drawn
\begin{lstlisting}
t     = 0
alpha = 0
# update the objects
while t < 2*TM:
	# set the frame rate (you can choose a higher rate to accelerate the program)
	rate(100)
	# update the drawn trajectory with the current position
	M.trajectory.append(pos=M.pos)
	# update the velocity and position
	M.pos , M.velocity = evolve_mercury(M.pos , M.velocity , alpha)
	# update time
	t = t + dt
\end{lstlisting}

If the starting values are chosen as advised in the previous section, then the students should end up with a trajectory as depicted in figure \ref{fig:MercuryOrbit-a0-small}.
At this point one might ask students to vary $\Delta t$ and observe its effects on the orbit [see also figure \ref{fig:MercuryOrbit-a0-small-dt-large}].
To get the perihelion motion, the additional force term described in equation (\ref{eq:newton_art}) has to be "turned on" for $\alpha\neq0$ [figure \ref{fig:MercuryOrbit-a6-small}].
This is a good opportunity to let the students play with the size of $\alpha$ and get a feeling for its impact on the trajectories.
However, it might be advisable to not start out with the correct mercury values but a more excentric trajectory by choosing e.g. $r_{MS}(0)=6R_0$ [figure \ref{fig:MercuryOrbit-a6-big}].
Then the perihelion motion is better visible.
\begin{figure}[htb]
	\centering
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/a0T5dt20.png}
		\caption{\label{fig:MercuryOrbit-a0-small}}
	\end{subfigure}
	~
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/num-err.png}
		\caption{\label{fig:MercuryOrbit-a0-small-dt-large}}
	\end{subfigure}
	~
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/a6T5dt20.png}
		\caption{\label{fig:MercuryOrbit-a6-small}}
	\end{subfigure}
	~
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/a6T10dt20.png}
		\caption{\label{fig:MercuryOrbit-a6-big}}
	\end{subfigure}
	\captionsetup{singlelinecheck=off}
	\caption[]{\label{fig:MercuryOrbit}
		Different mercury orbits for:
		\begin{itemize}
		\item[(a)] $\alpha=0$ and $\Delta t = \Delta t_0  / 20$;
		\item[(b)] $\alpha=0$ and $\Delta t = \Delta t_0  \times 2$, where the black line represents the orbit of (a) and the red line is the orbit for the larger time steps;
		\item[(c)] $\alpha=10^{6}$ and $\Delta t = \Delta t_0 / 20$;
		\item[(d)] $\alpha=10^{6}$ and $\Delta t = \Delta t_0 / 20$ but $r_{MS}(0) = 6 R_0$.
		\end{itemize}
		Here, the time steps are defined by $\Delta t_0 \equiv 2 v_{M}(0) / c_a$ and the images are screenshots of the simulation (for modified colors).
		Note that for different starting values of $r_{MS}(0)$, also the numbers of days for a "mercury year" changes.
	}
\end{figure}

@@

%\textit{What exactly do we want the students / advise the teachers to do? I think a nice approach would be to let the students implement the code to extract the perihelion motion and make a list of some values of alpha and the corresponding angles. Then discuss, that $\alpha=1$ is the natural size for $\alpha$ and that to extract the value of angle they need to exploit the linear dependence. Finally they can think on whether the value they get is reasonable and the teacher can explain the rest of section? What do you think?}\\

To calculate the expected size of the perihelion motion for a modified gravitational force, the students need to
\begin{itemize}
\item extract multiple positions of the perihelion $\vec r_{MS} ( t_\mathrm{ph}^{(n)} )$ for a fixed value of $\alpha$ and simulations times which covers several orbits;
\item calculate the angle between the perihelions for two successive turns and compute the average angle $\delta \Theta (\alpha)$ over all individually computed angles;
\item repeat the above points for different values of $\alpha$ to confirm the linear dependence of $\delta \Theta (\alpha)$;
\item interpolate $\delta \Theta (\alpha)$ to $\alpha=1$ using that $\delta \Theta (0) = 0$ (modulo numerical errors).
\end{itemize}

@@

\begin{figure}[htb]
	\centering
	\includegraphics{figs/alpha-angle.pdf}
	\caption{\label{fig:AlphaAngle} Linear relation between $\alpha$ and the perihelion motion $\delta \Theta$ for $\Delta t=2v_M(0)/c_a/100$.}
\end{figure}

To extract the perihelion motion at $\alpha = 1$, we use the fact that $\delta \Theta$ is zero for $\alpha =0$ --- if one could completely eliminate numerical errors.
Considering one result from the numeric simulation, e.g. $\delta\Theta(\alpha = 10^6) \approx 10.0^{\degree}$, the sought-after angle can be calculated by a linear interpolation as
\begin{equation}
	\delta\Theta (\alpha=1) = \frac{10.0^{\degree}}{10^6} = 0.036''
	\, .
\end{equation}
Depending on the knowledge of the students, multiple points and also estimated numerical uncertainties could be used to extract this value by a linear regression.
Note that, when comparing the results of the simulation with the experimental result, one Earth year corresponds to $T_E/T_M \approx 365d/88d\approx 4.15$ mercury years.
Therefore, the perihelion motion per 100 earth years is $0.036''\cdot 415\approx 15''$.
To test whether this is a reasonable result, the students could first follow the dimensional analysis and then compare to the actual value for mercury as outlined above.

@@

\begin{figure}[htb]
	\centering
	\includegraphics[width=.99\textwidth]{figs/angular-variaton.pdf}
	\caption{\label{fcc3}Motion of perihelion $\delta\Theta$ in degrees depending on the number of turns $N_t$ for different time steps $\Delta t$ (color) and different values for $\alpha$ (columns).
	Here, $\Delta t_0 \equiv 2 v_M(0)/c_a$.
	Offsets of $\delta \Theta(\alpha=0)$ from zero can be viewed as an error magnitude estimated and oscillations in $\delta \Theta$ indicate to coarse time steps.
}
\end{figure}

@@

The idea of dimensional analysis is that in a system that can be controlled by expanding the relevant quantities (like the force) in some small parameter(s), the coefficients in the expansion should turn out to be of order unity (that means anything between about 0.1 and 10 is fine - but 0.01 or 100 is irritating) --- parameters in line with this are called "natural".
Applied to the problem at hand given by Eq.~(\ref{eq:newton}) this statement implies that from naturalness one would expect that the parameter $\alpha$ is of order 1 when one uses for $r$ the average distance Mercury-sun, namely $\bar r=6\times 10^7$ km.
From this one estimates for the expected angular shift per orbit
\begin{equation}
\delta \Theta \simeq 2\pi\left(\frac{r_S}{\bar r}\right) = (\pi \times 10^{-7}) \ \mbox{rad} = (2\times 10^{-5})^o = (7\times 10^{-2}) ^{''} \ ,
\end{equation}
which leads to a shift of about $30^{''}$ in 100 earth years to be compared to the empirical value of $43^{''}$.
Thus indeed the amount of perihelion motion of Mercury is in line with expectations, $if$ --- and this is an important "if" --- the Newtonian dynamics is simply the leading term of some more general underlying theory.
In particular, no new scales enter in the correction terms.

>>>>

To allow for a movement of the perihelion we need to modify the force that appears in Eq.~(\ref{eq:newton}).
This modification should be such that it depends on $r$ and it vanishes for $r\to \infty$.
Based on the discussion presented above we may multiply the right hand side with a factor
$(1+\alpha(r_S/r))$, where $\alpha$ denotes some dimensionless parameter. 
However, besides $r_S$, which is the characteristic length scale for the gravitational potential
of the sun, one can define one more parameter, which is characteristic for the dynamics of mercury, namely
\begin{equation}
a^2=\frac{\vec L\,^2}{m^2c^2}= \frac{(\vec r\times \dot{\vec r} \, )^2}{c^2} \ ,
\end{equation}
where $\vec L$ denotes the angular momentum, which is a constant of motion for central potentials.
It is easily verified that $a^2$ carries dimensions of a length squared.  Therefore we may use  the following ansatz
for the modified potential
\begin{equation}
\ddot{\vec r} = - \frac{c^2}{2}\left(\frac{r_S}{r^2}\right)\left(1+\alpha\frac{r_S}{r}+\beta\frac{a^2}{r^2}\right) \, \frac{\vec{r}}{r} \ .
\label{eq:newton_art}
\end{equation}
For the system at hand $a^2$ may be estimated from the parameters of mercury at its perihelion ---  $\abs{\vec v_M(0)} = 59$km/s and 
$ r_{MS}=\abs{\vec r_{MS}(0)} = 46 \cdot 10^6$km~\cite{MercuryFactSheet}, where both $\vec r$ and $\dot{\vec r}$ are perpendicular to each other. 
We thus find that $(a^2/ r_{MS}^2) = 4\cdot10^{-8}$. Furthermore $(r_S/r_{MS})=6\cdot 10^{-8}$. 
Therefore one may expect from both correction terms an effect similar in size. Furthermore, any term higher order in either $a$ or $r_S$
should be suppressed by additional six orders of magnitude and thus can be safely neglected.
The parameters $\alpha$ and $\beta$ could now either be extracted from a fit to data or be calculated from the underlying
theory --- here General relativity with the result
\begin{equation}\label{eq:art-prediction}
	\alpha = 0 \, , \quad \beta = 3 \, .
\end{equation}

However, since we have only a single number to study, the angular motion of the perihelion,
we can only fix either $\alpha$ or $\beta$.

@@

In \python{}, this reads
\begin{lstlisting}
# Definition of parameters
rM0 = 4.6    # in units of R0
vM0 = 0.51   # in units of R0/T0
c_a = 1.01   # in units of R0/T0**2
TM  = 88.    # in units of T0
rS  = 3.0e-7 # in units of R0
a2  = 8.2e-7 # in units of R0**2
\end{lstlisting}
So far we only fixed the length of the vectors.
Next, we set up the initial directions, which will describe the motion in the two-dimensional plane.  We will build on the existing \python{} module \texttt{VPython}, which provides an implementation for treating vectors as well as their visualization.
The first object of interest is a \texttt{vector}, which takes three-dimensional coordinates as its input.
Because of our choice of initial conditions (we picked the initial vectors in the perihelion), the velocity of mercury is perpendicular to the vector which connects mercury and sun (see figure \ref{fig:sun_merc}):
\begin{lstlisting}
# Import the class vector from vpythons module visual
from vpython import vector
# Initialize distance and velocity vectors
vec_rM0 = vector(0, rM0, 0)
vec_vM0 = vector(vM0, 0, 0)
\end{lstlisting}
According to Eq.~(\ref{eq:newton}), the force which acts on mercury changes the velocity of mercury which eventually changes the position.
However, we first need to fix the time step $\Delta t$ and use the estimate of Eq.~(\ref{eq:check}) as a guidance
\begin{lstlisting}
# Definition of the time step
dt = 2 * vM0 / c_a / 20
\end{lstlisting}
Here the factor $1/20$ makes sure that $\Delta t$ is indeed consistent with Eq.~(\ref{eq:check}).
Now we are in the position to calculate location and velocity of the planet at $t_0+\Delta t$ using the following equations\footnote{%
	Note, that there are different notions for numerically integrating differential equations with different accuracys.
	Thus, the ordering and exact equations for updating $a_{MS}$, $v_M$ and $r_M$ are not unique, but all correct prescriptions lead to the same result for sufficiently small $\Delta t$.
	For this particular problem, the following definition is in good balance between numerically stable and inexpensive to compute.
}%
\begin{lstlisting}
# Compute the strength of the acceleration
temp = 1 + alpha * rS / vec_rM_old.mag + beta * a2 / vec_rM_old.mag**2
aMS  = c_a * temp / vec_rM_old.mag**2
# Multiply by the direction
vec_aMS = - aMS * ( vec_rM_old / vec_rM_old.mag )
# Update velocity vector
vec_vM_new = vec_vM_old + vec_aMS * dt
# Update position vector
vec_rM_new = vec_rM_old + vec_vM_new * dt
\end{lstlisting}
Note basic vector operations are already implemented in the predefined \texttt{vector} class.
The difference and sum of two vectors, or the scalar vector multiplication return vectors themselves.
Also the magnitude of a vector --- \texttt{vector.mag} --- is a property of the vector and can be easily extracted.

It is handy to use \texttt{Python}s \texttt{function}s to embed repeating structures ("DRY" --- Don't Repeat Yourself)
\begin{lstlisting}
# Define the coordinate and velocity update function
def evolve_mercury(vec_rM_old, vec_vM_old, alpha, beta):
	# Compute the strength of the acceleration
	temp = 1 + alpha * rS / vec_rM_old.mag + beta * a2 / vec_rM_old.mag**2
	aMS  = c_a * temp / vec_rM_old.mag**2
	# Multiply by the direction
	vec_aMS = - aMS * ( vec_rM_old / vec_rM_old.mag )
	# Update velocity vector
	vec_vM_new = vec_vM_old + vec_aMS * dt
	# Update position vector
	return vec_rM_new, vec_vM_new

# Call the function
vec_rM_new, vec_vM_new = evolve_mercury(vec_rM_old, vec_vM_old, 0.0, 0.0)
\end{lstlisting}
\python{}s syntax enforces a clean programming style: it is necessary that the body of the function is indented (by an arbitrary but consistent amount of \texttt{space}s or \texttt{tab}s) relative to the definition statement of the function.
Furthermore, \python{} is an Interpreter language.
Each line of the code is executed when the Interpreter passes it.
For this reason, all the variables defined before the function are global and thus known to the function.
The variables within the function (e.g. \texttt{aMS}, \texttt{vec\_aMS}, $\dots$) are local and do not exist beyond the scope of the function.

Finally, we can describe the evolution by a \texttt{while}-loop
\begin{lstlisting}%[float,floatplacement=H]
t     = 0.0
alpha = 0.0
beta  = 0.0
# Execute the loop as long as t < 2*TM
while t < 2*TM:
	vec_rM, vec_vM = evolve_mercury(vec_rM, vec_vM, alpha, beta)
	t = t + dt
\end{lstlisting}
where for the start we set the parameter $\alpha=0 = \beta$ in order first study the properties of the pure 
$1/r^2$ force.
Note the required indent of the loop structure similar to the indent of a function.
In each iteration of the \texttt{while}-loop, the previous distance and velocity are used to compute the new values --- which directly overwrites the previous values and thus enter the next iteration.
The total runtime \texttt{2*TM} is the amount of "virtual" days the simulations should run.
To describe at least one full orbital period, this time needs to be larger than $T_M$.
With the previous choice for \texttt{dt}, this corresponds to roughly $N_T \approx 2 \cdot 10^3$ evolution steps.
Note that the exact time it takes for mercury complete a full revolution depends on the initial coordinates and velocities as well as the accuracy of the computation.
One might encourage students to analyze this in the beginning.

@@

Last but not least, instead of updating the object-unrelated vectors in the while loop, we update the object-related vectors which will eventually be drawn
\begin{lstlisting}
t     = 0.0
alpha = 0.0
beta  = 0.0
# update the objects
while t < 2*TM:
	# set the frame rate (you can choose a higher rate to accelerate the program)
	rate(100)
	# update the drawn trajectory with the current position
	M.trajectory.append(pos=M.pos)
	# update the velocity and position
	M.pos , M.velocity = evolve_mercury(M.pos , M.velocity , alpha, beta)
	# update time
	t = t + dt
\end{lstlisting}

If the starting values are chosen as advised in the previous section, then the students should end up with a trajectory as depicted in figure \ref{fig:MercuryOrbit-a0-small}.
At this point one might ask students to vary $\Delta t$ and observe its effects on the orbit [see also figure \ref{fig:MercuryOrbit-a0-small-dt-large}].
To get the perihelion motion, the additional force term described in equation (\ref{eq:newton_art}) has to be "turned on" for $\alpha\neq0$ [figure \ref{fig:MercuryOrbit-a6-small}].
This is a good opportunity to let the students play with the size of $\alpha$ and get a feeling for its impact on the trajectories.
However, it might be advisable to not start out with the correct mercury values but a more excentric trajectory by choosing e.g. $r_{MS}(0)=6R_0$ [figure \ref{fig:MercuryOrbit-a6-big}].
Then the perihelion motion is better visible.
\begin{figure}[htb]
	\centering
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/a0T5dt20.png}
		\caption{\label{fig:MercuryOrbit-a0-small}}
	\end{subfigure}
	~
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/num-err.png}
		\caption{\label{fig:MercuryOrbit-a0-small-dt-large}}
	\end{subfigure}
	~
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/a6T5dt20.png}
		\caption{\label{fig:MercuryOrbit-a6-small}}
	\end{subfigure}
	~
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/a6T10dt20.png}
		\caption{\label{fig:MercuryOrbit-a6-big}}
	\end{subfigure}
	\captionsetup{singlelinecheck=off}
	\caption[]{\label{fig:MercuryOrbit}
		Different mercury orbits for $\beta=0$ and
		\begin{itemize}
		\item[(a)] $\alpha=0$ and $\Delta t = \Delta t_0  / 20$;
		\item[(b)] $\alpha=0$ and $\Delta t = \Delta t_0  \times 2$, where the black line represents the orbit of (a) and the red line is the orbit for the larger time steps;
		\item[(c)] $\alpha=10^{6}$ and $\Delta t = \Delta t_0 / 20$;
		\item[(d)] $\alpha=10^{6}$ and $\Delta t = \Delta t_0 / 20$ but $r_{MS}(0) = 6 R_0$.
		\end{itemize}
		Here, the time steps are defined by $\Delta t_0 \equiv 2 v_{M}(0) / c_a$ and the images are screenshots of the simulation (for modified colors).
		Note that for different starting values of $r_{MS}(0)$, also the numbers of days for a "mercury year" changes.
	}
\end{figure}

@@

%\textit{What exactly do we want the students / advise the teachers to do? I think a nice approach would be to let the students implement the code to extract the perihelion motion and make a list of some values of alpha and the corresponding angles. Then discuss, that $\alpha=1$ is the natural size for $\alpha$ and that to extract the value of angle they need to exploit the linear dependence. Finally they can think on whether the value they get is reasonable and the teacher can explain the rest of section? What do you think?}\\
To extract the linear dependence of the perihelion motion depending on $\alpha$ and $\beta$, we recommend to set one parameter to zero and vary the other.
Both parameter have similar effects on the perihelion motion.
As pointed out in Eq. \ref{eq:art-prediction}, general relativity predicts $(\alpha=0, \beta=3)$.
To calculate the expected size of the perihelion motion for a modified gravitational force, the students need to
\begin{itemize}
\item extract multiple positions of the perihelion $\vec r_{MS} ( t_\mathrm{ph}^{(n)} )$ for a fixed value of $\alpha$ and $\beta$ and simulations times which covers several orbits;
\item calculate the angle between the perihelions for two successive turns and compute the average angle $\delta \Theta (\alpha)$ over all individually computed angles;
\item repeat the above points for different values of $\alpha$ and $\beta$ to confirm the linear dependence of $\delta \Theta (\alpha, \beta)$;
\item interpolate $\delta \Theta (\alpha, \beta)$ to $(\alpha=1,\beta=0)$ or $(\alpha=0,\beta=3)$ using that $\delta \Theta (0,0) = 0$ (modulo numerical errors).
\end{itemize}

@@

\begin{figure}[htb]
	\centering
	\begin{subfigure}[c]{0.49\textwidth}
		\includegraphics[width=\textwidth]{figs/alpha-angle.pdf}
		\caption{\label{fig:angle-alpha}}
	\end{subfigure}
	\begin{subfigure}[c]{0.49\textwidth}
		\includegraphics[width=\textwidth]{figs/beta-angle.pdf}
		\caption{\label{fig:beta-alpha}}
	\end{subfigure}
	\caption{\label{fig:AlphaAngle} Linear relation between $\alpha$ [$\beta$] and and the perihelion motion $\delta \Theta$ for $\Delta t=2v_M(0)/c_a/100$ and $\beta=0$ [$\alpha=0$].}
\end{figure}

To extract the perihelion motion at $\alpha = 0$ and $\beta = 3$, we use the fact that $\delta \Theta$ is zero for $\alpha =0 = \beta$ --- if one could completely eliminate numerical errors.
Considering one result from the numeric simulation, e.g., $\delta\Theta(\alpha = 0, \beta=10^6) \approx 10.7^{\degree}$, the sought-after angle can be calculated by a linear interpolation as
\begin{equation}
	\delta\Theta (\alpha=0, \beta=3) = \frac{10.6^{\degree}}{10^6} \times 3 = 0.116''
	\, .
\end{equation}
Depending on the knowledge of the students, multiple points and also estimated numerical uncertainties could be used to extract this value by a linear regression.
Note that, when comparing the results of the simulation with the experimental result, one Earth year corresponds to $T_E/T_M \approx 365d/88d\approx 4.15$ mercury years.
Therefore, the perihelion motion per 100 earth years is $0.116''\cdot 415\approx 48.2''$.
This is close to experimental value of $\delta\Theta_{\mathrm{exp}} = 43.0''$.

@@

\begin{figure}[htb]
	\centering
	\includegraphics[width=.99\textwidth]{figs/angular-variaton.pdf}
	\caption{\label{fcc3}Motion of perihelion $\delta\Theta$ in degrees depending on the number of turns $N_t$ for different time steps $\Delta t$ (color) and different values for $\alpha$ (columns).
	Here, $\Delta t_0 \equiv 2 v_M(0)/c_a$.
	Offsets of $\delta \Theta(\alpha=0, \beta=0)$ from zero can be viewed as an error magnitude estimated and oscillations in $\delta \Theta$ indicate to coarse time steps.
}
\end{figure}

@@

The idea of dimensional analysis is that in a system that can be controlled by expanding the relevant quantities (like the force) in some small parameter(s), the coefficients in the expansion should turn out to be of order unity (that means anything between about 0.1 and 10 is fine - but 0.01 or 100 is irritating) --- parameters in line with this are called "natural".
Applied to the problem at hand given by Eq.~(\ref{eq:newton}) this statement implies that from naturalness one would expect that the parameter $\alpha$ is of order 1 when one uses for $r$ the average distance Mercury-sun, namely $\bar r=6\times 10^7$ km.
From this one estimates for the expected angular shift per orbit
\begin{equation}
\delta \Theta \simeq 2\pi\left(\frac{r_S}{\bar r}\right) = (\pi \times 10^{-7}) \ \mbox{rad} = (2\cdot 10^{-5})^o = (7\cdot 10^{-2}) ^{''} \ ,
\end{equation}
or
\begin{equation}
\delta \Theta \simeq 2\pi\left(\frac{a^2}{\bar r^2}\right) = (\pi \times 5 \cdot 10^{-8}) \ \mbox{rad} = (\cdot 10^{-5})^o = (4\cdot 10^{-2}) ^{''} \ ,
\end{equation}
which leads to a shift of about $30^{''}$ [$15^{''}$] in 100 earth years to be compared to the empirical value of $43^{''}$.
Thus indeed the amount of perihelion motion of Mercury is in line with expectations, $if$ --- and this is an important "if" --- the Newtonian dynamics is simply the leading term of some more general underlying theory.
In particular, no new scales enter in the correction terms.
